---
title: "deconv_indiv_slide_qc"
output: 
  html_document:
    df_output: paged
    toc: TRUE
    theme: 'simplex'

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = 'hide')
```

```{r}

suppressPackageStartupMessages({
  library(Seurat)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(patchwork)
library(ComplexHeatmap)
library(circlize)
library(gridExtra) 
library(ggdendro)
  library(readr)
library(data.table)
library(DT)
  library(fgsea)
  library(SpaceFold)
})

# load visium helper functions
source("/data/peer/sam/treg_dep/scripts/visium/visium_helper_functions.R")
# figure directory
fig_dir = "/data/peer/sam/treg_dep/visium/results/deconv_qc"

```

```{r}
# read in the visium data, now a processed RDS file
d.m = readRDS( '/data/peer/sam/treg_dep/visium/data/requant/processed/treg_dep_visium_merged_quant2_20220829.rds')
name_conv_df = read.csv( "/data/peer/sam/treg_dep/visium/bayesprism/viter1_nameConvSheet.csv", h = T, stringsAsFactors = F)
# make specific mods
name_conv_df[name_conv_df$make_name == 'Basophil','orig_name'] <- 'Basophil/Mast'
name_conv_df[name_conv_df$make_name == 'Platelet','orig_name'] <- 'Platelet/Mega'
name_conv_df[name_conv_df$make_name == 'Car4..capillary','orig_name'] <- 'aCap'
name_conv_df[name_conv_df$make_name == 'Plvap..capillary','orig_name'] <- 'gCap'

# read in cell type fractions
## these will be available in same github directory
bp_data_folder = "/data/peer/sam/treg_dep/visium/bayesprism/deconv_viter1_indiv/"

ct_frac_files = list.files(bp_data_folder, pattern = "*_theta.csv", full.names = T)
ct_cv_files = list.files(bp_data_folder, pattern = "*_thetaCV.csv", full.names = T)

# read these in as a list with the same names as d, the 10x data
ct_frac = lapply(ct_frac_files, read.csv, row.names = 1)
names(ct_frac) = gsub("_theta.csv", "", basename(ct_frac_files))

# fix rownames to match d.m
for(i in names(ct_frac)){
  rownames(ct_frac[[i]]) <- paste(i, rownames(ct_frac[[i]]), sep = "_")
}

# read these in as a list with the same names as d, the 10x data
ct_cv = lapply(ct_cv_files, read.csv, row.names = 1)
names(ct_cv) = gsub("_thetaCV.csv", "", basename(ct_cv_files))

# fix rownames to match d.m
for(i in names(ct_cv)){
  rownames(ct_cv[[i]]) <- paste(i, rownames(ct_cv[[i]]), sep = "_")
}

# read in positive spots
ssm = readRDS(file.path(bp_data_folder, 'sf_selected_spot_matrix.rds'))
# fix rownames to match
for(x in names(ssm)){
  rownames(ssm[[x]]) = paste(x, rownames(ssm[[x]]), sep = "_")
}
ssm.m = do.call(rbind, ssm)
colnames(ssm.m) = make.names(colnames(ssm.m))
# filter for only spots in the other ones

```


```{r}

# melt ct fraction values and filter for only spots that are in d.m
ct_frac.m = do.call(rbind, ct_frac)
rownames(ct_frac.m) = gsub("^.*\\.", "", rownames(ct_frac.m))
ct_frac.m = ct_frac.m[rownames(ct_frac.m) %in% colnames(d.m),]
section_id = gsub("_[AGCT]+\\-[0-9]*$", "", rownames(ct_frac.m))

ct_frac.sum = ct_frac.m %>% 
  mutate(section_id = section_id) %>%
  group_by(section_id) %>%
  summarise_all(mean)
  
```


# cell fraction comparison across serial sections

Visualize for each serial section comparison. 

```{r}

ct_frac.sum.sec = ct_frac.sum %>% reshape2::melt() %>%
  tidyr::pivot_wider(names_from = section_id, values_from = value)

#pdf(file.path(fig_dir, 'indiv_avgThetaComparison.pdf'), height = 6, width = 6)
# plot each of the 4 individually
(
  (ggplot(ct_frac.sum.sec, aes(x = Ctrl1_A1, y = Ctrl1_B1)) +
  geom_point() +
  theme_classic() +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = 'grey') + 
    theme(axis.text = element_text(size = 12), 
          axis.title = element_text(size = 14)) +
    labs(x = 'Control 1 section 1', 
         y = 'Control 1 section 2')) + 
  (ggplot(ct_frac.sum.sec, aes(x = Ctrl2_C1, y = Ctrl2_D1)) +
  geom_point() +
  theme_classic() +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = 'grey') + 
    theme(axis.text = element_text(size = 12), 
          axis.title = element_text(size = 14)) +
    labs(x = 'Control 2 section 1', 
         y = 'Control 2 section 2'))
  ) /
  ((ggplot(ct_frac.sum.sec, aes(x = DT1_A1, y = DT1_B1)) +
  geom_point() +
  theme_classic() +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = 'grey') + 
    theme(axis.text = element_text(size = 12), 
          axis.title = element_text(size = 14)) +
    labs(x = 'Treg depleted 1 section 1', 
         y = 'Treg depleted 1 section 2')) + 
    (ggplot(ct_frac.sum.sec, aes(x = DT2_C1, y = DT2_D1)) +
  geom_point() +
  theme_classic() +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = 'grey') + 
    theme(axis.text = element_text(size = 12), 
          axis.title = element_text(size = 14)) +
    labs(x = 'Treg depleted 2 section 1', 
         y = 'Treg depleted 2 section 2'))
  )
#dev.off()
```
Or visualize all together. 

```{r}

ct_frac.sum.stack = rbind(setNames(ct_frac.sum.sec[,c(2,3)], c('section_1', 'section_2')), 
                          setNames(ct_frac.sum.sec[,c(4,5)], c('section_1', 'section_2')), 
                          setNames(ct_frac.sum.sec[,c(6,7)], c('section_1', 'section_2')), 
                          setNames(ct_frac.sum.sec[,c(8,9)], c('section_1', 'section_2')))

#pdf(file.path(fig_dir, 'indiv_avgThetaComparison_allTogether.pdf'), height = 4, width = 4)
ggplot(ct_frac.sum.stack, aes(x = section_1, y = section_2)) +
  geom_point() +
  theme_classic() +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = 'grey') + 
  labs(x = 'Serial section 1\naverage cell fraction', y = 'Serial section 2\naverage cell fraction') + 
    theme(axis.text = element_text(size = 12), 
          axis.title = element_text(size = 14)) +
  ggpubr::stat_cor(method = "spearman", label.x = 0, label.y = .3, size = 4)
#dev.off()
```

# theta CV values for cell types

Looking at the coefficient of variation of theta values during bayesprism deconvolution. 

```{r}

# melt ct fraction values and filter for only spots that are in d.m
ct_cv.m = do.call(rbind, ct_cv)
rownames(ct_cv.m) = gsub("^.*\\.", "", rownames(ct_cv.m))
ct_cv.m = ct_cv.m[rownames(ct_cv.m) %in% colnames(d.m),]
section_id = gsub("_[AGCT]+\\-[0-9]*$", "", rownames(ct_cv.m))

# filter the positive spot matrix to contain the same spots in the same order
ssm.m = ssm.m[rownames(ct_cv.m),]

# extract positive spots for each cell type
ct_cv.pos = extractPositiveSpots(ct_cv.m, ssm.m)
ct_cv.pos$celltype = left_join(ct_cv.pos, name_conv_df, by = c('celltype' = 'make_name')) %>%
  .$orig_name
```

```{r}

ggplot(ct_cv.pos, aes(x = celltype, y = value)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5), 
        axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16)) +
  labs(x = "Cell type", y = "Cell fraction C.V.")

```


```{r}
# look at the 90th percentile for each of the cell types
ct_cv.q9 = ct_cv.pos %>% 
  group_by(celltype) %>%
  summarise(q90 = quantile(value, .9))
  
```

Plot the 90th percentile of cv values for each cell type in positive spots. 

```{r}

#pdf(file.path(fig_dir, '90th_percentile_CV_indiv.pdf'), height = 3, width = 3)
ggplot(ct_cv.q9, aes(x = 'All celltypes', y = q90)) +
  geom_boxplot(width = .6) + 
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16), 
        axis.title.x = element_blank()) +
  labs(y = '90th percentile C.V.') + 
  ylim(c(0,1))
#dev.off()
```



# robustness to downsample

```{r}

# read in the cell fractions from all the runs
ds_folder = '/data/peer/sam/treg_dep/visium/bayesprism/deconv_viter1_downsample'
ds_files = list.files(ds_folder, "*_theta.csv", recursive = T, full.names = T)


# read these in
ct_frac_ds = lapply(ds_files, read.csv, row.names = 1)
names(ct_frac_ds) <- unlist(lapply(strsplit(ds_files, "/"), "[[", 9))
ct_frac_comp = ct_frac$DT2_C1
rownames(ct_frac_comp) = gsub("DT2_C1_", "", rownames(ct_frac_comp))
# filter all of the downsampled runs for spots that are in the comparison section
ct_frac_ds = lapply(ct_frac_ds, function(x){
  x[rownames(x) %in% rownames(ct_frac_comp),]
})
ct_frac_comp = ct_frac_comp[rownames(ct_frac_ds$run_1),]
```

For each downsampled run, what is the pearson correlation of the cell type fractions for each cell type in each spot?

This should be done setting all cell fractions below .001 (our cutoff) to 0 to avoid comparisons that are not

```{r}

pcds = do.call(rbind, lapply(names(ct_frac_ds), function(x){
  #print(head(x))
  ct_cor = unlist(lapply(colnames(ct_frac_ds[[x]]), function(ct){
    ct_frac_comp_ct = ct_frac_comp[,ct]
    ct_frac_comp_ds = ct_frac_ds[[x]][,ct]
    cor(ct_frac_comp_ct, ct_frac_comp_ds, method = 'spearman') 
  }
  ))
  data.frame(spearman_cor = ct_cor, 
                     celltype = colnames(ct_frac_ds[[x]]), 
             run = x)
})
)
  
pcds$celltype = left_join(pcds, name_conv_df, by = c('celltype' = 'make_name')) %>%
  .$orig_name
```

```{r, fig.height=2, fig.width=3}
pdf(file.path(fig_dir, 'spearman_cor_in_downsampled.pdf'), height = 5, width = 6)
ggplot(pcds, aes(x = celltype, y = spearman_cor)) +
  geom_boxplot() +
  #geom_point() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5), 
        axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16)) +
  labs(x = "Cell type", y = "Spearman correlation to \nnon-down sampled")
  #ylim(c(.5, 1))

ggplot(pcds, aes(x = celltype, y = spearman_cor)) +
  geom_boxplot() +
  #geom_point() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5), 
        axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16)) +
  labs(x = "Cell type", y = "Spearman correlation to \nnon-down sampled") +
  ylim(c(0,1))
dev.off()
```

```{r}

#pdf(file.path(fig_dir, 'spearman_cor_in_downsampled_dotplot.pdf'), height = 5, width = 6)
# make this a dotplot with std error bar
pcds %>%
  group_by(celltype) %>%
  summarise(mean_spearman = mean(spearman_cor), 
            upper = max(spearman_cor), 
            lower = min(spearman_cor)) %>%
  ggplot(., aes(x = celltype, y = mean_spearman)) +
  #geom_boxplot() +
  geom_errorbar(aes(ymin = lower, ymax = upper),col = "grey60", 
                width =  .75) +
  geom_point(size = 1) +
  
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5), 
        axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16)) +
  labs(x = "Cell type", y = "Spearman correlation to \nnon-down sampled") +
  ylim(c(0,1))
#dev.off()

```

Write out this data for future needs

```{r, eval = F, include = F}

write.csv(pcds, file.path(fig_dir, 'spearman_cor_in_downsampled_df.csv'), 
          quote = F, row.names = F)


```


