---
title: "EMF factor usage"
output: 
  html_document:
    df_output: paged
    toc: TRUE
    theme: 'simplex'

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = 'hide')
```

```{r}
suppressPackageStartupMessages({
  library(Seurat)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(patchwork)
library(ComplexHeatmap)
library(circlize)
library(gridExtra) 
library(ggdendro)
  library(readr)
library(data.table)
library(DT)
  library(fgsea)
  library(sf)
  library(limma)
})

# load my functions
source("/data/peer/sam/treg_dep/scripts/visium/visium_helper_functions.R")
# figure directory
fig_dir = "/data/peer/sam/treg_dep/visium/results/factor_usage"

```


# Factor expression niche analysis

```{r, message = F, results = 'hide'}

## load 10x data and cell fractions

# load data and remove off tissue points
# this folder will be set to where 10x visium data files are stored
base_folder = '/data/peer/sam/treg_dep/visium/data/requant'

# set working dir
#setwd(base_folder)

# dirs where visium samples are
data_folders <- list.dirs(base_folder, recursive = T)
data_folders <- data_folders[grepl("outs$", data_folders)]

# read in visium data
d <- lapply(data_folders, Load10X_Spatial)
names(d) <- unlist(lapply(strsplit(data_folders, "/"), "[[", 10)) %>% gsub("_IGO_11440_[0-9]+", "", .)

# add necessary metadata to the tissue sections
lapply(names(d), function(x){
  d[[x]]@meta.data <<- mutate(d[[x]]@meta.data, section = x, 
                              treatment = ifelse(grepl("Ctrl", x), 'Ctrl', 'DT'))
})
# read in cell type fractions
## these will be available in same github directory
bp_data_folder = "/data/peer/sam/treg_dep/visium/bayesprism/deconv_viter1"

ct_frac_files = list.files(bp_data_folder, pattern = "*_theta.csv", full.names = T)

# read these in as a list with the same names as d, the 10x data
ct_frac = lapply(ct_frac_files, read.csv, row.names = 1)
names(ct_frac) = gsub("_theta.csv", "", basename(ct_frac_files))



# filter the 10x objects that I have for only spots that are in deconvolution
d = lapply(names(d), function(x){
  subset(d[[x]], cells = rownames(ct_frac[[x]]))
})
names(d) <- unlist(lapply(strsplit(data_folders, "/"), "[[", 10)) %>% gsub("_IGO_11440_[0-9]+", "", .)

# read in selected spot matrix
ssm = read.csv(file.path(bp_data_folder, 'selected_spot_matrix.median3mix.csv' ), 
               row.names = 1, stringsAsFactors = F)

# read in cell type name conversion
name_conv_df = read.csv( "/data/peer/sam/treg_dep/visium/bayesprism/viter1_nameConvSheet.csv", h = T, stringsAsFactors = F)

```


## Shared and unique genes on factors

Find shared genes involved in IFN and IC response across lineages as well as specific genes induced in these factors in particular lineages. 

```{r}

# function to get the genes along each factor
get_factor_genes = function(fac, gene_names){
  fac_mean = mean(fac)
  fac_std = sd(fac)
  
  fac_thresh = fac_mean + (2*fac_std)
  
  # get the factor genes and order by their score
  fac_scores = fac[fac > fac_thresh]
  names(fac_scores) = gene_names[fac > fac_thresh] %>% stringr::str_to_title()
  fac_genes = sort(fac_scores, decreasing = T) #%>% names()
  
  # return the gene names of associated terms to factor
  return(fac_genes)
}

read_factor_files = function(file_list){
  fac.l = invisible(lapply(file_list, readr::read_csv, show_col_types = F))
  names(fac.l) = sapply(file_list, basename) %>% gsub("*_factor_|\\.csv", "", .)
  
  # bind into matrix
  fac.mat = fac.l %>% purrr::reduce(left_join, by = '...1') %>% as.data.frame()
  rownames(fac.mat) = fac.mat$...1
  fac.mat = fac.mat[,-1]
  
  return(fac.mat)
}



```

```{r}

# list of lists for the factor file names
mm_factor_files = list(endo = list.files(path = "/data/peer/sam/treg_dep/mouse/results/schpf/endo", pattern = "*.csv", full.names = T), 
                       fib = list.files(path = "/data/peer/sam/treg_dep/mouse/results/schpf/fib", pattern = "*.csv", full.names = T), 
                       myeloid = list.files(path = "/data/peer/sam/treg_dep/mouse/results/schpf/myeloid", pattern = "*.csv", full.names = T))

mm_factor.l = lapply(mm_factor_files, read_factor_files)

fac_genes = lapply(mm_factor.l, function(x){
  apply(x, 2, get_factor_genes, rownames(x))
})
```

### factor genes in deconvolved

Check for the number of genes on each of the factors of interest that are shared in the visium and within the deconvolved expression values. 

```{r, include = F, eval = F}
deconv_genes = ct_exp$Ctrl1_A1_Ctrl1_B1$Pericyte %>% colnames()
gene_counts = readRDS(file = '/data/peer/sam/treg_dep/visium/data/requant/processed/gene_counts_visium_merged_quant2_20220829.rds')
gene_exp = names(gene_counts)[gene_counts >= 10]

# endo
fac_of_interest_endo = c( 'Factor-3', 'Factor-8', 'Factor-12', 'Factor-15', 'Factor-17', 'Factor-19')
endo_dc_ov = do.call(rbind, lapply(fac_of_interest_endo, function(gl){
  data.frame(factor_check = gl, 
             total = length(fac_genes$endo[[gl]]),
             overlap = sum(names(fac_genes$endo[[gl]]) %in% deconv_genes)) %>%
    mutate(frac_overlap = overlap / total)
})
)
endo_nodc_ov = do.call(rbind, lapply(fac_of_interest_endo, function(gl){
  data.frame(factor_check = gl, 
             total = length(fac_genes$endo[[gl]]),
             overlap = sum(names(fac_genes$endo[[gl]]) %in% gene_exp)) %>%
    mutate(frac_overlap = overlap / total)
})
)

```


This analysis supports using undeconvolved as many genes are missed in the deconvolved. 


### compute factor scores

```{r}

# compute scores for each of the gene sets

# first combine and normlize the visium data from serial section 1
# create a merged object for different downstream analyses
d.m = merge(d[[1]], d[2:length(d)], add.cell.ids = names(d))

d.m_rep1 = d.m[,d.m@meta.data$section %in% c('Ctrl1_A1', 'Ctrl2_C1', 'DT1_A1', 'DT2_C1')]
# get genes to filter out
gene_sum = rowSums(d.m_rep1@assays$Spatial@counts)
gene_rm = gene_sum < 10
print(sum(gene_rm))
# SCTransform normalize all spots in the merged object for first technical replicate and then subset these for the tumor lesion response
DefaultAssay(d.m_rep1) <- 'Spatial'
d.m_rep1 = SCTransform(d.m_rep1, assay = "Spatial", verbose = F)
# fix the images so I don't have all of the ones I don't need
d.m_rep1@images = d.m_rep1@images[c('slice1', 'slice1.2', 'slice1.4', 'slice1.6')]

```

```{r}

# calculate scores
# I am also going to include the growth factor in this
score_genes_list = list(fac_genes$endo$`Factor-3`, fac_genes$endo$`Factor-8`, fac_genes$endo$`Factor-12`, fac_genes$endo$`Factor-15`, fac_genes$endo$`Factor-17`, fac_genes$endo$`Factor-19`, 
                        fac_genes$fib$`Factor-17`, fac_genes$fib$`Factor-11`, fac_genes$fib$`Factor-14`, fac_genes$fib$`Factor-9`, fac_genes$fib$`Factor-22`, fac_genes$fib$`Factor-21`, 
                        fac_genes$myeloid$`Factor-2`, fac_genes$myeloid$`Factor-5`, fac_genes$myeloid$`Factor-13`, fac_genes$myeloid$`Factor-15`, fac_genes$myeloid$`Factor-17`, fac_genes$myeloid$`Factor-21`, fac_genes$myeloid$`Factor-23`)
names(score_genes_list) = c('endo inflammatory capillary', 'endo angiogenesis',  'endo growth factor', 'endo inflammation/hypoxia', 'endo inflammation EndMT', 'endo IFN', 'fib MSC-like', 'fib VEGFC', 'fib ECM', 'fib IFN', 'fib inflammatory cytokine', 'fib inflammation/MT', 'myl C1Q+ macrophage','myl monocyte coagulation', 'myl Csf3r monocyte', 'myl proliferation', 'myl IFN', 'myl monocyte hypoxia', 'myl Arg1+ macrophage' )
# convert to names
score_genes_list = lapply(score_genes_list, names)


d.m_rep1 = AddModuleScore(d.m_rep1, score_genes_list, name = names(score_genes_list))

# fix the names in the metadata
colnames(d.m_rep1@meta.data)[grepl('^endo|^fib|^myl', colnames(d.m_rep1@meta.data))] = gsub("[0-9]+$", "", colnames(d.m_rep1@meta.data)[grepl('^endo|^fib|^myl', colnames(d.m_rep1@meta.data))])

```



```{r}
fac_m = d.m_rep1@meta.data[,grepl('^endo|^fib|^myl', colnames(d.m_rep1@meta.data))] %>% as.matrix()
# compute pearson correlation
fac_cor = cor(fac_m)

```

```{r}

# save file so I don't need to recompute every time
write.csv(fac_m, file.path(fig_dir, 'foctor_of_interest_moduleScores_rep1.csv'), quote = F)


```

```{r, fig.height=6, fig.width=8}

Heatmap(fac_cor)


```


```{r}

# melt and join to tumor information
fac_m.m = fac_m %>% as.matrix() %>% reshape2::melt() %>%
  left_join(data.frame(Var1 = rownames(d.m_rep1@meta.data), 
                      # tumor_spot = d.m_rep1@meta.data$Tumor_spot, 
                       condition = d.m_rep1@meta.data$treatment), 
            by = 'Var1') %>%
  mutate(lineage = ifelse(grepl("endo", Var2), 'Endothelial', ifelse(grepl("fib", Var2), 'Fibroblast', 'Myeloid'))) %>%
  dplyr::rename(barcode = Var1, scRNA_factor = Var2, factor_score = value)

```

### cell type specific usage

Get spots where particular cell types are detected and compare factor scores. 


```{r}

# loop through  each cell type of interest and extract spots where it is present
# for each relevant factor of interest in that cell type, compute the mean, sd, sem
cell_interest_factors = list(
  'Pericyte' = c('fib.IFN'),
  "Col14a1..fibroblast" = c('fib.MSC.like', 'fib.VEGFC', 'fib.ECM', 'fib.IFN', 'fib.inflammatory.cytokine', 'fib.inflammation.MT'), 
  "Myofibroblast" = c('fib.MSC.like', 'fib.VEGFC', 'fib.ECM', 'fib.IFN',  'fib.inflammation.MT'), 
  "Col13a1..fibroblast" = c( 'fib.IFN', 'fib.inflammatory.cytokine', 'fib.inflammation.MT'),
  "MSC" = c('fib.MSC.like'),
  "Monocyte" = c('myl.monocyte.coagulation', 'myl.Csf3r.monocyte', 'myl.IFN', 'myl.monocyte.hypoxia'), 
  "Alveolar.macrophage" = c('myl.proliferation', 'myl.IFN'), 
  "Macrophage" = c('myl.C1Q..macrophage', 'myl.Arg1..macrophage', 'myl.IFN'), 
  "Car4..capillary" = c('endo.inflammatory.capillary', 'endo.IFN'), 
  "Plvap..capillary" = c('endo.angiogenesis', 'endo.growth.factor', 'endo.inflammation.hypoxia', 'endo.inflammation.EndMT', 'endo.IFN' ),
  "Lymphatic.endothelial" = c('endo.inflammation.hypoxia', 'endo.inflammation.EndMT')
)


cell_fac_diff = do.call(rbind, lapply(names(cell_interest_factors), function(ct){
  pos_spots = rownames(ssm[ssm[,ct],])
  fac_f = dplyr::filter(fac_m.m, barcode %in% pos_spots & scRNA_factor %in% cell_interest_factors[[ct]])
  do.call(rbind, lapply(cell_interest_factors[[ct]], function(fac_int){
    fac_test = dplyr::filter(fac_f, scRNA_factor == fac_int)
    t.test(factor_score ~ condition, data = fac_test) %>% broom::tidy() %>%
      mutate(celltype = ct, 
             scRNA_factor = fac_int)
  }))
  
  

})
)

cell_fac_diff$p.adj = p.adjust(cell_fac_diff$p.value, method = "bonferroni")

```

Barplot representing this test. 

```{r}

ggplot(cell_fac_diff, aes(x = scRNA_factor, y = estimate * -1)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  facet_wrap(~ celltype, scales = "free") +
  theme_classic()


```

```{r}
# fix the names
cell_fac_diff$celltype = left_join(cell_fac_diff, name_conv_df, by = c('celltype' = 'make_name')) %>% .$orig_name

fac_conv_df = data.frame(make_name = unique(cell_fac_diff$scRNA_factor), 
                         orig_name = c('fib IFN', 'fib MSC-like', 'fib VEGFC', 'fib ECM',  'fib inflammatory cytokine', 'fib inflammation/MT', 
                                       'myl monocyte coagulation','myl Csf3r monocyte', 'myl IFN', 'myl monocyte hypoxia','myl proliferation', 'myl C1Q+ macrophage',     'myl Arg1+ macrophage', 
                                       'endo inflammatory capillary', 'endo IFN',  'endo angiogenesis',  'endo growth factor', 'endo inflammation/hypoxia', 'endo inflammation EndMT'))
cell_fac_diff$scRNA_factor_mod = left_join(cell_fac_diff, fac_conv_df, by = c('scRNA_factor' = 'make_name')) %>% .$orig_name #%>% 
  #gsub("endo |fib |myl ", "", .)
# arrange in proper order for thing
cell_fac_diff = do.call(rbind, list(
  cell_fac_diff[grepl("myl", cell_fac_diff$scRNA_factor),],
  cell_fac_diff[grepl("fib", cell_fac_diff$scRNA_factor),],
  cell_fac_diff[grepl("endo", cell_fac_diff$scRNA_factor),]
)) %>% mutate(
  celltype = forcats::fct_relevel(factor(celltype), as.character(unique(celltype))), 
         scRNA_factor = forcats::fct_relevel(factor(scRNA_factor), as.character(unique(scRNA_factor)))) %>%
  mutate(scRNA_factor_mod = forcats::fct_relevel(factor(scRNA_factor_mod), as.character(unique(scRNA_factor_mod))))
```

```{r}

# pdf(file.path(fig_dir, 'factor_t_across_condition_dotplot.pdf'),
#   height = 6, width = 8)
cell_fac_diff %>%
  mutate(p.adj = ifelse(p.adj < 1e-215, 1e-216, p.adj)) %>%
  mutate(estimate = ifelse(p.adj > .001, NA_integer_, estimate)#, 
         #celltype = forcats::fct_relevel(factor(celltype), as.character(unique(celltype))), 
         #scRNA_factor = forcats::fct_relevel(factor(scRNA_factor), as.character(unique(scRNA_factor)))
         ) %>%
  #arrange(scRNA_factor_mod) %>%
 # mutate(scRNA_factor_mod = forcats::fct_relevel(factor(scRNA_factor_mod), as.character(unique(scRNA_factor_mod)))) %>%
  
  ggplot(., aes(x = celltype, y = scRNA_factor_mod)) +
  geom_point(aes(color = estimate * -1,
                 size = -log10(p.adj))#, color = 'black', pch = 21
             ) +
  theme_few() +
  
  scale_color_gradient2( limits = c(-.07, .07), high = scales::muted('red'), low = scales::muted('blue'), 
                        oob = scales::squish, 
                        name = 'Treg depleted - Control\nmean factor score') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5), 
        axis.text = element_text(size = 14), 
        axis.title = element_blank()) +
  scale_y_discrete( labels = gsub("endo |fib |myl ", "", unique(cell_fac_diff$scRNA_factor_mod)))
#dev.off()



```

```{r, eval = F, include = F}
# write out stats
write.csv(cell_fac_diff, file = file.path(fig_dir, 'factor_condition_ttest_20220910.csv'), 
          row.names = F, quote =)



```



