---
title: "IFN_IC_niche_v5"
output: 
  html_document:
    df_output: paged
    toc: TRUE
    theme: 'simplex'

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = 'hide')
```

```{r}
suppressPackageStartupMessages({
  library(Seurat)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(patchwork)
library(ComplexHeatmap)
library(circlize)
library(gridExtra) 
library(ggdendro)
  library(readr)
library(data.table)
library(DT)
  library(fgsea)
  library(sf)
  library(limma)
})

# load my functions
source("/data/peer/sam/treg_dep/scripts/visium/visium_helper_functions.R")
# figure directory
fig_dir = "/data/peer/sam/treg_dep/visium/results/factor_usage/"

# set seed for reproducibility
set.seed(1234)

```


# Factor expression niche analysis

```{r, message = F, results = 'hide'}

## load 10x data and cell fractions

# load data and remove off tissue points
# this folder will be set to where 10x visium data files are stored
base_folder = '/data/peer/sam/treg_dep/visium/data/requant'

# set working dir
#setwd(base_folder)

# dirs where visium samples are
data_folders <- list.dirs(base_folder, recursive = T)
data_folders <- data_folders[grepl("outs$", data_folders)]

# read in visium data
d <- lapply(data_folders, Load10X_Spatial)
names(d) <- unlist(lapply(strsplit(data_folders, "/"), "[[", 10)) %>% gsub("_IGO_11440_[0-9]+", "", .)

# add necessary metadata to the tissue sections
lapply(names(d), function(x){
  d[[x]]@meta.data <<- mutate(d[[x]]@meta.data, section = x, 
                              treatment = ifelse(grepl("Ctrl", x), 'Ctrl', 'DT'))
})
# read in cell type fractions
## these will be available in same github directory
bp_data_folder = "/data/peer/sam/treg_dep/visium/bayesprism/deconv_viter1"

ct_frac_files = list.files(bp_data_folder, pattern = "*_theta.csv", full.names = T)

deconv_base_folder = '/data/peer/sam/treg_dep/visium/bayesprism'
deconv_folders = c('deconv_viter1')

# read in cell fraction files and collapse to a single data frame
ct_frac = lapply(deconv_folders, function(folder){
  ct_frac_files = list.files(file.path(deconv_base_folder, folder), pattern = "*_theta.csv", full.names = T)
  ct.l = lapply(ct_frac_files, read.csv, row.names = 1)
  #names(ct.l) = gsub("_theta.csv", "", ct_frac_files)
  ct.lm = do.call(rbind,ct.l)
  colnames(ct.lm) = make.names(colnames(ct.lm))
  return(ct.lm)
})
names(ct_frac) = deconv_folders

# first split the cellfractions into the different tissue sections

split.idx = remove_barcode(rownames(ct_frac$deconv_viter1))
ct_frac.l = split(ct_frac$deconv_viter1, factor(split.idx))


# change the rownames of the slides to be indexed by section as well 
invisible(lapply(names(d), function(x){
#  rownames(d[[x]]@meta.data) <<- paste(d[[x]]@meta.data$section, rownames(d[[x]]@meta.data), sep = ".")
  d[[x]] <<- RenameCells(d[[x]], new.names = paste(d[[x]]@meta.data$section, rownames(d[[x]]@meta.data), sep = "_"))
}))

# filter the 10x objects that I have for only spots that are in deconvolution
d = lapply(names(d), function(x){
  subset(d[[x]], cells = rownames(ct_frac.l[[x]]))
})
names(d) <- unlist(lapply(strsplit(data_folders, "/"), "[[", 10)) %>% gsub("_IGO_11440_[0-9]+", "", .)



# add cell type fractions to each section
lapply(names(d), function(x){
  ct_obj = CreateAssayObject(counts = ct_frac.l[[x]] %>% t())
  d[[x]][['BP']] <<- ct_obj
})



# read in selected spot matrix
ssm = read.csv(file.path(bp_data_folder, 'selected_spot_matrix.median.csv' ), 
               row.names = 1, stringsAsFactors = F)

# read in cell type name conversion
name_conv_df = read.csv( "/data/peer/sam/treg_dep/visium/bayesprism/viter1_nameConvSheet.csv", h = T, stringsAsFactors = F)

```


## Shared and unique genes on factors

Find shared genes involved in IFN and IC response across lineages as well as specific genes induced in these factors in particular lineages. 

```{r}

# function to get the genes along each factor
get_factor_genes = function(fac, gene_names){
  fac_mean = mean(fac)
  fac_std = sd(fac)
  
  fac_thresh = fac_mean + (2*fac_std)
  
  # get the factor genes and order by their score
  fac_scores = fac[fac > fac_thresh]
  names(fac_scores) = gene_names[fac > fac_thresh] %>% stringr::str_to_title()
  fac_genes = sort(fac_scores, decreasing = T) #%>% names()
  
  # return the gene names of associated terms to factor
  return(fac_genes)
}

read_factor_files = function(file_list){
  fac.l = invisible(lapply(file_list, readr::read_csv, show_col_types = F))
  names(fac.l) = sapply(file_list, basename) %>% gsub("*_factor_|\\.csv", "", .)
  
  # bind into matrix
  fac.mat = fac.l %>% purrr::reduce(left_join, by = '...1') %>% as.data.frame()
  rownames(fac.mat) = fac.mat$...1
  fac.mat = fac.mat[,-1]
  
  return(fac.mat)
}



```

```{r}

# list of lists for the factor file names
mm_factor_files = list(endo = list.files(path = "/data/peer/sam/treg_dep/mouse/results/schpf/endo", pattern = "*.csv", full.names = T), 
                       fib = list.files(path = "/data/peer/sam/treg_dep/mouse/results/schpf/fib", pattern = "*.csv", full.names = T), 
                       myeloid = list.files(path = "/data/peer/sam/treg_dep/mouse/results/schpf/myeloid", pattern = "*.csv", full.names = T))

mm_factor.l = lapply(mm_factor_files, read_factor_files)

fac_genes = lapply(mm_factor.l, function(x){
  apply(x, 2, get_factor_genes, rownames(x))
})
```
```{r, eval = F, inclue = F}
# code to output genes selected for each factor

fac_genes.m = do.call(rbind, lapply(fac_genes, function(x) { reshape2::melt(lapply(x, function(y){ as.matrix(y)})
                                                                                   )})) %>%
  tibble::rownames_to_column(var = 'lineage') %>%
  mutate(lineage = gsub("\\.[0-9]+$", "", lineage)) %>%
  dplyr::rename(gene = Var1, gene_score = value, factor = L1) %>%
  dplyr::select(-Var2)

# write this out in a processed data location
write.csv(fac_genes.m, file = '/data/peer/sam/treg_dep/mouse/results/schpf/processed/mouse_factor_associatedGenes.csv', 
          quote = F, row.names = F)

```


```{r}

# get genes that are shared and unique for particular factors
ifn_shared = Reduce(intersect, list(names(fac_genes$endo$`Factor-19`), names(fac_genes$fib$`Factor-9`), names(fac_genes$myeloid$`Factor-17`)))
ifn_unique = list()
ifn_unique[['endo']] = setdiff(names(fac_genes$endo$`Factor-19`), names(fac_genes$fib$`Factor-9`)) %>% setdiff(., names(fac_genes$myeloid$`Factor-17`))
ifn_unique[['fib']] = setdiff(names(fac_genes$fib$`Factor-9`), names(fac_genes$endo$`Factor-19`)) %>% setdiff(., names(fac_genes$myeloid$`Factor-17`))
ifn_unique[['myeloid']] = setdiff(names(fac_genes$myeloid$`Factor-17`), names(fac_genes$endo$`Factor-19`)) %>% setdiff(., names(fac_genes$fib$`Factor-9`) )

# get genes that are shared and unique for particular factors
ic_shared = Reduce(intersect, list(names(fac_genes$endo$`Factor-15`), names(fac_genes$fib$`Factor-22`), names(fac_genes$myeloid$`Factor-21`)))
ic_unique = list()
ic_unique[['endo']] = setdiff(names(fac_genes$endo$`Factor-15`), names(fac_genes$fib$`Factor-22`)) %>% setdiff(., names(fac_genes$myeloid$`Factor-21`))
ic_unique[['fib']] = setdiff(names(fac_genes$fib$`Factor-22`), names(fac_genes$endo$`Factor-15`)) %>% setdiff(., names(fac_genes$myeloid$`Factor-21`))
ic_unique[['myeloid']] = setdiff(names(fac_genes$myeloid$`Factor-21`), names(fac_genes$endo$`Factor-15`)) %>% setdiff(., names(fac_genes$fib$`Factor-22`) )
```

How many genes are shared or unique out of total?

```{r}
print("IFN")
print(paste("There are ", length(ifn_shared), 'shared IFN genes across lineages out of', length(unique(c(names(fac_genes$endo$`Factor-19`), names(fac_genes$fib$`Factor-9`), names(fac_genes$myeloid$`Factor-17`)))), 'total genes'))
print(paste("endo unique count:", length(ifn_unique$endo)))
print(paste("fib unique count:", length(ifn_unique$fib)))
print(paste("myeloid unique count:", length(ifn_unique$myeloid)))

print("IC")
print(paste("There are ", length(ic_shared), 'shared IC genes across lineages out of', length(unique(c(names(fac_genes$endo$`Factor-15`), names(fac_genes$fib$`Factor-22`), names(fac_genes$myeloid$`Factor-21`)))), 'total genes'))
print(paste("endo unique count:", length(ic_unique$endo)))
print(paste("fib unique count:", length(ic_unique$fib)))
print(paste("myeloid unique count:", length(ic_unique$myeloid)))



```
```{r}
# shared genes
print("IC shared genes")
print(ic_shared)
print("IFN shared genes")
print(ifn_shared)



```

```{r}

# compute scores for each of the gene sets

# first combine and normlize the visium data from serial section 1
# create a merged object for different downstream analyses
d.m = merge(d[[1]], d[2:length(d)])#, add.cell.ids = names(d))
# get genes to filter out
gene_sum = rowSums(d.m@assays$Spatial@counts)
gene_rm = gene_sum < 10
# SCTransform normalize all spots in the merged object for first technical replicate and then subset these for the tumor lesion response
DefaultAssay(d.m) <- 'Spatial'
d.m_rep1 = SCTransform(d.m[!(gene_rm),d.m@meta.data$section %in% c('Ctrl1_A1', 'Ctrl2_C1', 'DT1_A1', 'DT2_C1')], assay = "Spatial", verbose = F)
# fix the images so I don't have all of the ones I don't need
d.m_rep1@images = d.m_rep1@images[c('slice1', 'slice1.2', 'slice1.4', 'slice1.6')]

```

```{r}

# calculate scores
# I am also going to include the growth factor in this
score_genes_list = list(ifn_shared, ifn_unique$endo, ifn_unique$fib, ifn_unique$myeloid, 
                        ic_shared, ic_unique$endo, ic_unique$fib, ic_unique$myeloid, 
                        fac_genes$endo$`Factor-12` %>% names())
names(score_genes_list) = c('ifn_shared', 'ifn_endo', 'ifn_fib', 'ifn_myeloid', 
                            'ic_shared', 'ic_endo', 'ic_fib', 'ic_myeloid', 
                            'gf_endo')
d.m_rep1 = AddModuleScore(d.m_rep1, score_genes_list, name = names(score_genes_list))

# fix the names in the metadata
colnames(d.m_rep1@meta.data)[grepl('ic_|ifn_|gf_', colnames(d.m_rep1@meta.data))] = gsub("[0-9]$", "", colnames(d.m_rep1@meta.data)[grepl('ic_|ifn_|gf_', colnames(d.m_rep1@meta.data))])

```

```{r}

# add metadata to the seurat visium objects to plot the factor scores in spatial context
score_dfs = split(d.m_rep1@meta.data[,names(score_genes_list)], d.m_rep1@meta.data$section)

for(i in names(score_dfs)){
  d[[i]]@meta.data = cbind(d[[i]]@meta.data, score_dfs[[i]])
}


```

```{r, eval = F, fig.height=8, fig.width=10}

# derive cutoffs for the max values to use
# to standardize the scales for each factor I will get the 95th percentile of all visium spots across all sections for each factor

feature_max = apply(d.m_rep1@meta.data[,names(score_genes_list)], 2, quantile, .99)
  
pdf(file.path(fig_dir, 'ifn_ic_geneScore_spatPlot.pdf'), 
    height = 8, width =)
invisible(lapply(d[c('Ctrl1_A1', 'Ctrl2_C1', 'DT1_A1', 'DT2_C1')], SpatialFeaturePlotList, names(score_genes_list), feature_max))

dev.off()

```
### write out the scores

```{r, eval = F}
saveRDS(score_dfs, file = file.path(fig_dir, 'ic_ifn_scores_rep1.csv'))
# write.csv(score_dfs, file.path(fig_dir, 'ic_ifn_scores_rep1.csv'), 
#           row.names = T, quote = F)

```

### gene loadings in shared vs unique

Show that the loadings of the shared genes in both settings are higher than the unique, justifying that these are providing the highest signal to noise of a similarly induced gene program. 

```{r}

# get scores
ifn_shared_score = list()
ifn_shared_score[['endo']] = fac_genes$endo$`Factor-19`[ifn_shared]
ifn_shared_score[['fib']] = fac_genes$fib$`Factor-9`[ifn_shared]
ifn_shared_score[['myeloid']] = fac_genes$myeloid$`Factor-17`[ifn_shared]


# get scores
ifn_unique_score = list()
ifn_unique_score[['endo']] = fac_genes$endo$`Factor-19`[ifn_unique$endo]
ifn_unique_score[['fib']] = fac_genes$fib$`Factor-9`[ifn_unique$fib]
ifn_unique_score[['myeloid']] = fac_genes$myeloid$`Factor-17`[ifn_unique$myeloid]


# get scores
ic_shared_score = list()
ic_shared_score[['endo']] = fac_genes$endo$`Factor-15`[ic_shared]
ic_shared_score[['fib']] = fac_genes$fib$`Factor-22`[ic_shared]
ic_shared_score[['myeloid']] = fac_genes$myeloid$`Factor-21`[ic_shared]

# get scores
ic_unique_score = list()
ic_unique_score[['endo']] = fac_genes$endo$`Factor-15`[ic_unique[['endo']]]
ic_unique_score[['fib']] = fac_genes$fib$`Factor-22`[ic_unique[['fib']]]
ic_unique_score[['myeloid']] = fac_genes$myeloid$`Factor-21`[ic_unique[['myeloid']]]

# process into single df
ic_ifn_score_df = rbind(
  do.call(rbind, list(lapply(ifn_shared_score, as.matrix) %>% reshape2::melt() %>%
                        mutate(gene_module = 'ifn'), 
                    lapply(ic_shared_score, as.matrix) %>% reshape2::melt() %>%
                        mutate(gene_module = 'ic')
                    )
          ) %>%
  mutate(type = 'shared'), 
    do.call(rbind, list(lapply(ifn_unique_score, as.matrix) %>% reshape2::melt() %>%
                        mutate(gene_module = 'ifn'), 
                    lapply(ic_unique_score, as.matrix) %>% reshape2::melt() %>%
                        mutate(gene_module = 'ic')
                    )
            ) %>%
  mutate(type = 'unique')
)


```

Plot the loadings of shared and unique genes in each lineage for each type signaling pathway. 

```{r}
# pdf(file.path(fig_dir, 'shared_vs_unique_geneScores.pdf'), 
#     height = 8, width = 12)
ggplot(ic_ifn_score_df, aes(x = type, y = value)) +
  geom_boxplot() +
  facet_wrap(gene_module ~ L1) +
  theme_classic() + 
  ylab("Gene score") +
  xlab("Gene type") + 
    theme(
        axis.text = element_text(size = 12), 
        axis.title = element_text(size = 14),
        strip.text.x = element_text(size = 14))
# dev.off()
```

This appears to be particularly true for the IFN, with IC less so but the trend still somewhat holding. 

### similar behavior of unique

```{r}


unique_cor = cor(d.m_rep1@meta.data[,c('ifn_shared', 'ifn_endo', 'ifn_fib', 'ifn_myeloid', 
                                       'ic_shared', 'ic_endo', 'ic_fib', 'ic_myeloid')], 
                 method = 'spearman')


```

```{r}
# pdf(file.path(fig_dir, 'shared_vs_unique_correlation.pdf'), 
#     height = 6, width = 8)
Heatmap(unique_cor, 
        name = "Spearman\ncorrelation")
#dev.off()

```

## define territories


Focus on IFN and IC, remove GF for now. 

```{r}

# need to add a constant to the distribution so that it will work
pseudo_count = 1

## IFN
ifn_fit = MASS::fitdistr(d.m_rep1@meta.data[,'ifn_shared'] + pseudo_count, 'gamma')

ifn_para <- ifn_fit$estimate
gamma_den = function(x, para) dgamma(x, shape = ifn_para[1], rate = ifn_para[2])
hist(d.m_rep1@meta.data[,'ifn_shared'] + 1, prob = TRUE)
curve(gamma_den, col = 2, add = TRUE)

## IC
ic_fit = MASS::fitdistr(d.m_rep1@meta.data[,'ic_shared'] + pseudo_count, 'gamma')

ic_para <- ic_fit$estimate
gamma_den = function(x, para) dgamma(x, shape = ic_para[1], rate = ic_para[2])
hist(d.m_rep1@meta.data[,'ic_shared'] + pseudo_count, prob = TRUE)
curve(gamma_den, col = 2, add = TRUE)


```

```{r}
# what is the cutoff that I should use
ifn_cutoff = qgamma(.01, shape =  ifn_para[1], rate = ifn_para[2], lower.tail = F) - pseudo_count
ic_cutoff = qgamma(.01, shape = ic_para[1], rate  = ic_para[2], lower.tail = F) - pseudo_count
#gf_cutoff = qgamma(.01, shape = gf_para[1], rate = gf_para[2], lower.tail = F) - pseudo_count

print(paste("IFN:", ifn_cutoff, "; IC:", ic_cutoff))#, "Growth factor:", gf_cutoff))
```

```{r}


ifn_niche = (d.m_rep1@meta.data[,'ifn_shared'] > ifn_cutoff) %>% as.factor() %>% as.numeric()
ic_niche = (d.m_rep1@meta.data[,'ic_shared'] > ic_cutoff) %>% as.factor() %>% as.numeric()
#gf_niche = (d.m_rep1@meta.data[,'gf_endo'] > gf_cutoff) %>% as.factor() %>% as.numeric()

#niche_labels = cbind(ifn_niche, ic_niche, gf_niche) - 1
niche_labels = cbind(ifn_niche, ic_niche) - 1
```

```{r}
# are there certain areas with multiple labels?
apply(niche_labels, 1, sum) %>% table()

```

There are 23 spots with an overlap. 

```{r}

niche_vec = apply(niche_labels, 1, function(x){
  #niche_names = c('IFN', 'IC', 'GF')
  niche_names = c('IFN', 'IC')
  paste(niche_names[as.logical(x)], collapse = "+")
}) 
niche_vec[niche_vec == ""] <- 'other'
table(niche_vec, useNA = 'ifany')
names(niche_vec) = rownames(d.m_rep1@meta.data)
```

```{r}

niche_df = niche_vec %>% as.matrix() %>% melt() %>%
  dplyr::select(barcode = Var1, niche = value)

# write out the niche labels
write.csv(niche_df, file = file.path(fig_dir, 'niche_labels_rep1.csv'), 
          row.names = F, quote = F)

# read if necessary
#niche_df = read.csv(file.path(fig_dir, 'niche_labels_rep1.csv'), stringsAsFactors = F)
```


```{r}

# add this metadata to the image objects to be plotted in the spatial context
invisible(lapply(names(d), function(x){
  #print(x)
  nl_filt = niche_vec[grepl(x, names(niche_vec))]
  #names(nl_filt) = gsub(paste0(x, '.'), "", names(nl_filt))
  print(paste(sum(names(nl_filt) %in% rownames(d[[x]]@meta.data)), "names out of", length(names(nl_filt))))
  #print(str(nl_filt))
  d[[x]]@meta.data[,'niche_label'] <<- nl_filt[rownames(d[[x]]@meta.data)]
  # d[[x]]@meta.data[,'PC1'] <<- pc_d[rownames(d[[x]]@meta.data), c('PC1')]
  # d[[x]]@meta.data[,'PC2'] <<- pc_d[rownames(d[[x]]@meta.data), c('PC2')]
  #d[[x]]@meta.data <<- cbind(d[[x]]@meta.data, pc_d[rownames(d[[x]]@meta.data), c('PC1','PC2')])
}))


```

### spatial plots of niches

```{r, fig.height = 10, fig.width=12}
niche_pal = c(ggthemes_data$tableau$`color-palettes`$regular$`Tableau 10`$value[c(2:3,6)], 'black')
#niche_pal = c(ggthemes_data$tableau$`color-palettes`$regular$`Tableau 10`$value[1:6], 'grey')
names(niche_pal) = c(#'GF', 
  'IFN', 'IC', 
                     #'IC+GF', 'IFN+GF', 
                     'IFN+IC', 'other')#, 'IFN+IC+GF', 'other')
#'IC+GF', 'IFN+GF', 'IFN+IC', 'IFN+IC+GF', 'other')


# pdf(file.path(fig_dir, 'ifn_ic_niche_spatPlot.pdf'), 
#     height = 10, width = 14)
SpatialPlotSections(d[c(1,3,5,7)], 'niche_label', scale_fill_manual(values = niche_pal, labels = names(niche_pal)))
#dev.off()
```






### enrichment vs normal tissue 

How I do this currently is to create an empirical null distribution by sampling without replacement from all spots across the tissue. Then, the mean cell type fraction for each cell type is compared to the means of each of the samples and the fraction of sample averages greater than the niche average is taken as the p value. 

```{r}

DefaultAssay(d.m) <- 'BP'
d.m_rep1.bp = subset(d.m, section %in% c('Ctrl1_A1', 'Ctrl2_C1', 'DT1_A1', 'DT2_C1'))
# using the cell_type enrichment function that I wrote in the visium_helper_functions script
mat.ct = d.m_rep1.bp@assays$BP@counts %>% t()
# I need to remove certain niches from the analysis because they only have 1 point
niche_enr = cell_type_enrichment(mat.ct, niche_vec, num_bg_samples = 10000, num_sample = 100)
# fix names
niche_enr$celltype = left_join(niche_enr, name_conv_df, by = c('celltype' = 'make_name')) %>% .$orig_name
```


```{r}

# plot the fold change in each niche and color by the empirical p value
# pdf(file.path(fig_dir, 'cell_enr_ifnIc_empP.pdf'),
#     height = 4, width = 8)
niche_enr %>% #dplyr::filter(celltype %in% (dplyr::filter(niche_enr, padj < .05) %>% .$celltype %>% unique())) %>%
  dplyr::filter(padj < .01) %>%
  mutate(padj = ifelse(padj > .01, NA_integer_, padj), 
         log2FC = ifelse(padj > .01, NA_integer_, log2FC)) %>%
  ggplot(., aes(x = celltype, y = log2FC)) +
  #geom_bar(aes(fill = num_gt / 10000), stat = 'identity') +
  geom_bar(aes(fill = padj), 
    stat = 'identity'#, fill = 'grey'
    ) +
  theme_few() +
  coord_flip() +
  facet_wrap(~ niche, ncol = 4, scales = "free_x") + 
  labs(y = 'Celltype RNA fraction log2 fold-change', 
       x = 'Celltype') + 
  scale_fill_continuous_tableau(palette = "Gray", name = 'Adjusted emprircal p', trans = 'reverse', 
                                limits = c(.01, 0)) +
  theme(legend.position = "bottom") + 
    theme( strip.text.x = element_text(size = 14),
        axis.text = element_text(size = 12), 
        axis.title = element_text(size = 14))
#dev.off()

```





